Model Checking Differentially Private Properties

### 2018-01-31 introduction.tex
fix short sentences

Designing differentially private mechanisms can be tedious for
sophisticated data analyses. Privacy leak has also been observed in
data analysis programs with differentially private
designs~\cite{M:12:SLSBDP,TKBW:17:PLAI}. This calls for formal
analysis of differential privacy on both designs and implementations.
In tis paper, we propose the logic $\dpCTLstar$ for specifying
differential privacy and investigate their model checking
problems. Data analysts can automatically verify their designs and
implementations with our techniques. Most interestingly, our
techniques can be adopted rather easily by existing probabilistic
model checkers. Privacy checking with existing tools can be attained
with minimal efforts and overheads. Hopefully, more interaction
between model checking and privacy analysis will follow.

@@

In order to illusrtrate applicability of our techniques, we give
detailed formalizations of several data analysis mechanisms in this
paper. 
In differential privacy, data analysis mechanisms are
but randomized algorithms. We follow the standard practice in
probabilistic model checking to formalize such mechanisms by
by Markov chains or Markov decision processes~\cite{Put05}. We give
formalizations of several differentially private mechanisms and
explain subtle privacy conditions.
When a data analysis mechanism does not
interact with its environment, it is formalized as a Markov
chain. Otherwise, its interactions are formalized by actions in
Markov decision processes. Our formalization effectively assumes that
actions are controlled by adversaries. It is subsequently necessary to
consider attacks from all possible adversaries in order to establish
differential privacy.
%\todo{What is a state?}

@@

We moreover extend the standard probabilistic model checking
algorithms to verify $\dpCTLstar$ properties automatically. For Markov
chains, states 
satisfying a subformula $\dpriv{\epsilon}{\delta} \phi$ are computed
by a simple variant of the model checking algorithm for Markov chains.
The time complexity of our algorithm is the same as
those of $\PCTLstar$ for Markov chains. The logic $\dpCTLstar$ obtains its
essentially expressiveness for free.
For Markov decision processes, we show that checking whether a state
satisfies $\dpriv{\epsilon}{\delta} \phi$ is undecidable.

@@

\paragraph{Contributions.} Our main contributions are threefold. (I)
We introduce the logic $\dpCTLstar$ for reasoning about properties in
differential privacy. We use the logic to express classical
differential privacy properties, and moreover, complex properties such as those
involving infinite behaviors can also be expressed. (II) We model several differential privacy mechanisms, including stream processing, in Markov chains or Markov decision processes.
(III) We show that the model checking 
problem for Markov chains is standard. For Markov decision processes we show that it is undecidable.

>>>>

Designing differentially private mechanisms can be tedious for
sophisticated data analyses. Privacy leak has also been observed in
data analysis programs with differentially private
designs~\cite{M:12:SLSBDP,TKBW:17:PLAI}. This calls for formal
analysis of differential privacy on both designs and implementations.
In tis paper, we propose the logic $\dpCTLstar$ for specifying
differential privacy and investigate their model checking
problems. Data analysts can automatically verify their designs and
implementations with our techniques. Most interestingly, our
techniques can be adopted easily by existing probabilistic
model checkers. Privacy checking with existing tools is attainable
with minimal efforts. More interaction
between model checking and privacy analysis hopefully will follow.

@@

In order to illusrtrate applicability of our techniques, we give
detailed formalizations of several data analysis mechanisms in this
paper. 
In differential privacy, data analysis mechanisms are
but randomized algorithms. We follow the standard practice in
probabilistic model checking to formalize such mechanisms by
by Markov chains or Markov decision processes~\cite{Put05}. We give
formalizations of several differentially private mechanisms and
explain subtle privacy conditions.
When a data analysis mechanism does not
interact with its environment, it is formalized as a Markov
chain. Otherwise, its interactions are formalized by actions in
Markov decision processes. Our formalization effectively assumes that
actions are controlled by adversaries. It is thus necessary to
consider all privacy attacks from adversaries in order to establish
differential privacy.
%\todo{What is a state?}

@@

We moreover extend the standard probabilistic model checking
algorithms to verify $\dpCTLstar$ properties automatically. For Markov
chains, states 
satisfying a subformula $\dpriv{\epsilon}{\delta} \phi$ are computed
by a simple variant of the model checking algorithm for Markov chains.
The time complexity of our algorithm is the same as
those of $\PCTLstar$ for Markov chains. The logic $\dpCTLstar$ obtains its
expressiveness essentially for free.
For Markov decision processes, checking whether a state
satisfies $\dpriv{\epsilon}{\delta} \phi$ is undecidable.

@@

\paragraph{Contributions.} Our main contributions are threefold. (I)
We introduce the logic $\dpCTLstar$ for reasoning about properties in
differential privacy. We use the logic to express classical
differential privacy properties, and moreover, complex properties such as those
involving infinite behaviors can also be expressed. (II) We model several differentially private mechanisms, including stream processing, in Markov chains or Markov decision processes.
(III) We show that the model checking 
problem for Markov chains is standard. For Markov decision processes we show that it is undecidable.

### 2018-01-31 main.tex
merge comments
adapted schedulers ect
fix short sentences

\paragraph{Organisation of the paper.}
Preliminaries are given in Section~\ref{section:preliminaries}. 
In Section~\ref{section:examples} we discuss how differential privacy mechanisms can be modeled in Markov chains.
The logic $\dpCTLstar$ and model checking algorithms for Markov chains are presented in
Section~\ref{section:dpCTL}. 
Section~\ref{section:specifying-properties} describes differential privacy properties using   $\dpCTLstar$.
They are followed by the extension to Markov decision processes and an
application in streaming process (Section~\ref{section:mdp}). Finally,
Section~\ref{section:conclusions} concludes our presentation.

@@

\subsection{MDPs for Streaming Processes}
\label{section:applications}
\input{applications}

\section{Conclusions}
\label{section:conclusions}
We have introduced $\dpCTLstar$ to reason about properties in differential privacy, and have investigated the model checking problem. For Markov chains, the model checking problem has polynomial complexity. The general MDP model checking problem is undecidable. We have discussed that some special cases are decidable. An interesting work is to identify more
decidable subclass. As an example, consider the formula $\bigwedge_{k \in \bbfN}
\dpriv{\epsilon}{\delta} (\X^k  00)$ which was used in specifying the pan privacy. For the case $\epsilon=\delta=0$, it reduces to a language equivalence problem for probabilistic automata. It is interesting to characterise other cases as well.
 Another interesting line of further works is to consider continuous perturbation (such as Laplace distribution used in \cite{DR:14:AFDP}) -- we would need Markov models with continuous state space for the modelling.

>>>>

\paragraph{Organisation of the paper.}
Preliminaries are given in Section~\ref{section:preliminaries}.
In Section~\ref{section:examples} we discuss how differentially private mechanisms can be modeled in Markov chains.
The logic $\dpCTLstar$ and model checking algorithms for Markov chains are presented in
Section~\ref{section:dpCTL}.
Section~\ref{section:specifying-properties} describes differential privacy properties using   $\dpCTLstar$.
They are followed by the extension to Markov decision processes and an
application in streaming process (Section~\ref{section:mdp}). Finally,
Section~\ref{section:conclusions} concludes our presentation.

@@

\subsection{MDPs for Stream Processing}
\label{section:applications}
\input{applications}

\section{Conclusions}
\label{section:conclusions}
We have introduced $\dpCTLstar$ to reason about properties in differential privacy, and have investigated the model checking problem. For Markov chains, the model checking problem has the same complexity as for $\PCTLstar$. The general MDP model checking problem is undecidable. We have discussed that some special cases are decidable. An interesting work is to identify more
decidable subclass. As an example, consider the formula $\bigwedge_{k \in \bbfN}
\dpriv{\epsilon}{\delta} (\X^k  00)$ which was used in specifying the pan privacy. For the case $\epsilon=\delta=0$, it reduces to a language equivalence problem for probabilistic automata. It is interesting to characterise other cases as well.
 Another interesting line of further works is to consider continuous perturbation (such as Laplace distribution used in \cite{DR:14:AFDP}) -- we would need Markov models with continuous state space for the modelling.
